---
title: "FLIM Analysis Template"
output: word_document
date: "2025-02-26"
author: "Sevde Coban"
---

```{r setup, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}
library(dplyr)
library(here)
library(stringr)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.heaight=12, dpi=330)
local_dest <- "/Users/scoban/Documents/Github/flim_pipeline/data"


# Import data
#dataframe is data imported using data_cleaning.R function
dataframe <- readRDS(file.path(local_dest, paste0("data_cleaning_outputnew_", ".RDS")))
# average_dataframe is a df of the mean G and mean fB
average_dataframe <- readRDS(file.path(local_dest, paste0("data_cleaning_outputnew_avg_", ".RDS")))

# Function to extract concentration from the 'Replicate_Group' column
extract_concentration <- function(dataframe, replicate_group_column) {
  get_concentration <- function(value) {
    if (grepl("untreated", value)) {
      return(0)
    }
    value <- gsub("STA-", "", value)
    value <- gsub("([0-9])-([0-9])", "\\1.\\2", value)
    value <- gsub("-", "", value)
    concentration <- gsub("uM", "", value)
  
    return(as.numeric(concentration))
  }
  
  dataframe$Concentration_uM <- sapply(dataframe[[replicate_group_column]], get_concentration)
  return(dataframe)
}


dataframe <- dataframe %>% extract_concentration('Replicate_Group')
average_dataframe <-average_dataframe %>% extract_concentration('Replicate_Group')

```

# Purpose

FLIM collects photon excitation information from cell images to examine metabolic changes within biologic samples. This information is Fourier transformed to collect two parameters: S-coordinate and G-coordinate. These parameters are used to calculate fraction bound (fB), which is the fraction of bound to unbound NADH within the cells we’re looking at. Fraction bound can tell you if the cells are using more glycolysis (more free NADH) or oxidative phosphorylation (more NADH is bound to an enzyme).

The purpose of this analysis is to compare the differences in fraction bound (fB) between STA treated and untreated samples.


# Z' Factor and % Cell Viability for Fraction Bound and G-Coordinates

```{r FB, echo=FALSE, warning=FALSE, message=FALSE}


# FB Z' FACTOR

# Three conditions: "untreated" "STA-1uM"   "STA-0-1uM"
# The low control is STA-0-1uM, high control is STA-1uM

dataframe <- dataframe %>%  
  group_by(Experiment) %>%
  mutate(cv_fb = sd(fB, na.rm = TRUE)/mean(fB, na.rm = TRUE)*100) %>%

  # Low control (STA-0-1uM)
  mutate(ctrl_0_mean_fb = mean(fB[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         ctrl_0_sd_fb = sd(fB[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_fb = fB - ctrl_0_mean_fb,

        #High control (STA-1uM)
         ctrl_100_mean_fb = mean(fB[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_minus_mean_fb = mean(signal_minus_ctrl_0_mean_fb[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_sd_fb = sd(fB[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         z_prime_factor_fb = 1 - (3*ctrl_100_sd_fb + 3*ctrl_0_sd_fb)/abs(ctrl_100_mean_fb - ctrl_0_mean_fb),
         plate_cv_fb = mean(cv_fb, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime <- dataframe$z_prime_factor_fb
plate_cv <- dataframe$plate_cv_fb

summary_dataframe <- data.frame(z_prime_factor = z_prime, percent_cv = plate_cv)
summary_dataframe <- summary_dataframe[1, , drop = FALSE]

kable(summary_dataframe, caption = "Fraction Bound Z'factor and %CV")
```


```{r mean_fb, echo=FALSE, warning=FALSE, message=FALSE}

# MEAN FB Z' FACTOR

average_dataframe <- average_dataframe %>%  
  group_by(Experiment) %>%
  mutate(cv_mean_fb_mean = sd(meanfB, na.rm = TRUE)/mean(meanfB, na.rm = TRUE)*100) %>%
  
  #Low control
  mutate(ctrl_0_mean_fb_mean = mean(meanfB[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         ctrl_0_sd_fb_mean = sd(meanfB[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_fb_mean = meanfB - ctrl_0_mean_fb_mean,
         #High control
         ctrl_100_mean_fb_mean = mean(meanfB[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_minus_mean_fb_mean = mean(signal_minus_ctrl_0_mean_fb_mean[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_sd_fb_mean = sd(meanfB[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         z_prime_factor_mean_fb = 1 - (3*ctrl_100_sd_fb_mean + 3*ctrl_0_sd_fb_mean)/abs(ctrl_100_mean_fb_mean - ctrl_0_mean_fb_mean),
         plate_cv_mean_fb = mean(cv_mean_fb_mean, na.rm = TRUE)) %>%
         ungroup()

z_prime_mean <- average_dataframe$z_prime_factor_mean_fb
plate_cv_mean <- average_dataframe$plate_cv_mean_fb


summary_df2 <- data.frame(z_prime_factor = z_prime_mean, percent_cv = plate_cv_mean)
summary_df2 <- summary_df2[1, , drop = FALSE]


library(knitr)

kable(summary_df2, caption = "Mean Fraction Bound Z'factor and %CV")


```
```{r gcoord, echo=FALSE, warning=FALSE, message=FALSE}

# G-COORDINATE Z' FACTOR


dataframe <- dataframe %>%  
  group_by(Experiment) %>% 
  mutate(cv_g = sd(G, na.rm = TRUE)/mean(G, na.rm = TRUE)*100) %>%
  
  #Low control
  mutate(ctrl_0_mean_g = mean(G[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         ctrl_0_sd_g = sd(G[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_g = G - ctrl_0_mean_g,
         #High control
         ctrl_100_mean_g = mean(G[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_minus_mean_g = mean(signal_minus_ctrl_0_mean_g[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_sd_g = sd(G[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         z_prime_factor_g = 1 - (3*ctrl_100_sd_g + 3*ctrl_0_sd_g)/abs(ctrl_100_mean_g - ctrl_0_mean_g),
         plate_cv_g = mean(cv_g, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime <- dataframe$z_prime_factor_g
plate_cv <- dataframe$plate_cv_g

summary_df3 <- data.frame(z_prime_factor = z_prime, percent_cv = plate_cv)
summary_df3 <- summary_df3[1, , drop = FALSE]

library(knitr)

kable(summary_df3, caption = "G-Coordinate Z'factor and %CV")


```

```{r mean gcoord, echo=FALSE, warning=FALSE, message=FALSE}

# MEAN G-COORDINATE Z' FACTOR


average_dataframe <- average_dataframe %>%  
  group_by(Experiment) %>%
  mutate(cv_mean_g_mean = sd(meanG, na.rm = TRUE)/mean(meanG, na.rm = TRUE)*100) %>%
  
  #Low control
  mutate(ctrl_0_mean_g_mean = mean(meanG[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         ctrl_0_sd_g_mean = sd(meanG[Replicate_Group == "STA-0-1uM"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_g_mean = meanG - ctrl_0_mean_g_mean,
         # High control
         ctrl_100_mean_g_mean = mean(meanG[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_minus_mean_g_mean = mean(signal_minus_ctrl_0_mean_g_mean[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         ctrl_100_sd_g_mean = sd(meanG[Replicate_Group == "STA-1uM"], na.rm = TRUE),
         z_prime_factor_mean_g = 1 - (3*ctrl_100_sd_g_mean + 3*ctrl_0_sd_g_mean)/abs(ctrl_100_mean_g_mean - ctrl_0_mean_g_mean),
         plate_cv_mean_g = mean(cv_mean_g_mean, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime_mean_g_coord <- average_dataframe$z_prime_factor_mean_g
plate_cv_mean_g_coord <- average_dataframe$plate_cv_mean_g


summary_df4 <- data.frame(z_prime_factor = z_prime_mean_g_coord, percent_cv = plate_cv_mean_g_coord)
summary_df4 <- summary_df4[1, , drop = FALSE]

library(knitr)

kable(summary_df4, caption = "Mean G-Coordinate Z'factor and %CV")


```

# Scatterplot of S- & G- Coordinates of STA Treatments


```{r plot1, echo=FALSE, warning=FALSE, message=FALSE,  fig.width=13, fig.height=10, dpi=330}

library(ggplot2)
library(viridis)



# previously colored by Plate)ID (org1) but this is only 1 plate.

scatterplot <- ggplot(dataframe, aes(x=G, y=S, color =Experiment)) + geom_point(size=4) + facet_wrap(~Replicate_Group + Concentration_uM) + labs(x = "G-coordinate", y = "S-coordinate", color = "Experiment") + scale_color_brewer(palette = "Dark2") + theme_bw(base_size=20) + ggtitle("S- & G- Coordinates") 


scatterplot


```




# Mixed Effects Models (Fraction Bound)


```{r boxplot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=9, dpi=300}

#setting up STA-1uM_df

library(dplyr)
library(ggplot2)
library(lme4)
library(ggbeeswarm)

df_all <- dataframe %>% mutate(Concentration = case_when(
  Replicate_Group == "untreated" ~ 0,
  Replicate_Group == "STA-0-1uM" ~ 0.1,
  Replicate_Group == "STA-1uM" ~ 1,
  TRUE ~ NA_real_
))


averages <- df_all %>%
  group_by(Experiment, Concentration) %>%
  summarize(mean_fB = mean(fB, na.rm=TRUE), .groups='drop')



box_superplot <- ggplot(df_all, aes(x = as.factor(Concentration), y = fB)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  geom_quasirandom(aes(color = as.factor(Experiment)), size = 2, alpha = 0.9) +
  geom_beeswarm(data = averages, aes(x = as.factor(Concentration), y = mean_fB,color = as.factor(Experiment)), size = 5)+ 
  scale_color_brewer(palette = "Dark2") +
  theme_bw(base_size=20) +
  ggtitle("Fraction Bound Across Conditions") +
  labs(x = "STA Concentration uM", y = "Fraction Bound", color = "Experiment")
  
  
box_superplot

```









```{r fixed, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=9, dpi=300}




# Fixed effects model (not considering experiment)
invisible(fixed_effects_model <- lm(fB ~ Concentration, data = df_all))


# Fixed effects plot
fixed_plot <- ggplot(data = df_all, aes(x = Concentration, y = fB)) +
    geom_point(alpha=0.9, color = "darkgray", size =3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    theme_bw(base_size=20) +
    ggtitle("STA Fraction Bound") + 
    labs(subtitle = "Fixed Effects Visualizer") +
    ylab("Fraction Bound") +
    xlab("STA Concentration (uM)")
  

fixed_plot

```







```{r mem, echo=FALSE, warning=FALSE, message=FALSE,fig.width=12, fig.height=9, dpi=300}
# Mixed effects model
invisible(mixed_effects_model <- lmer(fB ~ Concentration  + (1 | Experiment), data=df_all))


mixed_effects_plot_w_date <- ggplot(data = df_all, aes(x = Concentration, y = fB, color = as.factor(Experiment))) +
  geom_point(alpha=0.9, size=3) +
    geom_smooth(method = "lm", aes(group=Experiment), se = FALSE, size =1.2) +
  theme_bw(base_size=20) +
  scale_color_brewer(palette = "Dark2") +
    ggtitle("STA Fraction Bound") + 
    labs(subtitle = "Mixed Effects Visualizer") +
    labs(x = "STA Concentration (uM)", y = "Fraction Bound", color = "Experiment")


mixed_effects_plot_w_date



```

The mixed effect model examines the effects of STA Concentration on fraction bound, with experiment as a random variable to assess differences between experiment runs.

