---
title: "FLIM Analysis Template"
output: word_document
date: "2025-03-11"
author: "Sevde Coban"
---

```{r setup, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}

# Set up

library(dplyr)
library(here)
library(stringr)
library(knitr)
source("/Users/scoban/Documents/Github/danGoat/R/new_data_cleaner.R")

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.heaight=12, dpi=330)
local_dest <- "/Users/scoban/Documents/Github/flim_pipeline/data"


# Import data
#dataframe is data imported using data_cleaning.R function
dataframe <- readRDS(file.path(db_dest, paste0("data_cleaning_outputnew_", ".RDS")))
# average_dataframe is a df of the mean G and mean fB
average_dataframe <- readRDS(file.path(db_dest, paste0("data_cleaning_outputnew_avg_", ".RDS")))

```





```{r, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}
# Function to extract concentration from the 'Replicate_Group' column
extract_concentration <- function(dataframe, replicate_group_column = "Replicate_Group") {
  get_concentration <- function(value) {
    # Extract last part after underscore
    treatment <- sub(".*_", "", value)

    if (tolower(treatment) == "untreated") {
      return(0)
    }

    treatment <- gsub("STA-", "", treatment)
    treatment <- gsub("([0-9]+)-([0-9]+)", "\\1.\\2", treatment)  # 0-1 → 0.1
    treatment <- gsub("uM", "", treatment)

    numeric_value <- suppressWarnings(as.numeric(treatment))
    if (is.na(numeric_value)) {
      warning(paste("Could not extract concentration from:", value))
    }

    return(numeric_value)
  }

  dataframe$Concentration <- sapply(dataframe[[replicate_group_column]], get_concentration)
  return(dataframe)
}
```

# Purpose

FLIM collects photon excitation information from cell images to examine metabolic changes within biologic samples. This information is Fourier transformed to collect two parameters: S-coordinate and G-coordinate. These parameters are used to calculate fraction bound (fB), which is the fraction of bound to unbound NADH within the cells we’re looking at. Fraction bound can tell you if the cells are using more glycolysis (more free NADH) or oxidative phosphorylation (more NADH is bound to an enzyme).

The purpose of this analysis is to compare the differences in fraction bound (fB) between STA treated and untreated samples.


# Z' Factors for Fraction Bound and G-Coordinates
```{r, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}

# Function to calculate z' factor

calculate_zprime <- function(df, value_col, replicate_col = "Replicate_Group") {
  df_list <- list()

  for (exp in unique(df$Experiment)) {
    sub_df <- df[df$Experiment == exp, ]
    x <- sub_df[[value_col]]
    rep_group <- sub_df[[replicate_col]]

    # Use grepl to find control groups
    ctrl_0_vals <- x[grepl("STA-0-1uM", rep_group)]
    ctrl_100_vals <- x[grepl("STA-1uM", rep_group)]

    if (length(ctrl_0_vals) >= 2 && length(ctrl_100_vals) >= 2) {
      ctrl_0_mean <- mean(ctrl_0_vals, na.rm = TRUE)
      ctrl_0_sd <- sd(ctrl_0_vals, na.rm = TRUE)
      ctrl_100_mean <- mean(ctrl_100_vals, na.rm = TRUE)
      ctrl_100_sd <- sd(ctrl_100_vals, na.rm = TRUE)

      if (!is.na(ctrl_0_mean) && !is.na(ctrl_100_mean) && ctrl_0_mean != ctrl_100_mean) {
        z_prime <- 1 - (3 * ctrl_0_sd + 3 * ctrl_100_sd) / abs(ctrl_100_mean - ctrl_0_mean)
      } else {
        z_prime <- NA_real_
      }
    } else {
      z_prime <- NA_real_
    }

    df_list[[exp]] <- data.frame(Experiment = exp, z_prime_factor = z_prime)
  }

  return(do.call(rbind, df_list))
}
```




```{r, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}
# Extract concentrations
dataframe <- extract_concentration(dataframe)
average_dataframe <- extract_concentration(average_dataframe)

```



```{r, include=FALSE, echo=FALSE, message= FALSE, warning = FALSE}
# Finding z' values (mean and raw)
zprime_fb_raw  <- calculate_zprime(dataframe, "fB")
zprime_g_raw   <- calculate_zprime(dataframe, "G")
zprime_fb_avg  <- calculate_zprime(average_dataframe, "meanfB")
zprime_g_avg   <- calculate_zprime(average_dataframe, "meanG")


```


Here are the Z' factors for fB and G-Coordinate across the 3 experiments. Raw values are calculated using all individual measurements for each sample. Mean values first average indiividual measurements within each Sample_ID then calculates Z' factors using the averaged values. This minimizes intra-sample variability. 


```{r, echo=FALSE, message= FALSE, warning = FALSE}

# Summary table for word

summary_fb_raw <- data.frame(
  Experiment = unique(zprime_fb_raw$Experiment),
  z_prime_factor = zprime_fb_raw %>% select(Experiment, z_prime_factor_fb = z_prime_factor) %>% distinct() %>% pull(z_prime_factor_fb)
)

summary_fb_avg <- data.frame(
  Experiment = unique(zprime_fb_avg$Experiment),
  z_prime_factor = zprime_fb_avg %>% select(Experiment, z_prime_factor_mean_fb = z_prime_factor) %>% distinct() %>% pull(z_prime_factor_mean_fb)
)

summary_g_raw <- data.frame(
  Experiment = unique(zprime_g_raw$Experiment),
  z_prime_factor = zprime_g_raw %>% select(Experiment, z_prime_factor_g = z_prime_factor) %>% distinct() %>% pull(z_prime_factor_g)
)

summary_g_avg <- data.frame(
  Experiment = unique(zprime_g_avg$Experiment),
  z_prime_factor = zprime_g_avg %>% select(Experiment, z_prime_factor_mean_g = z_prime_factor) %>% distinct() %>% pull(z_prime_factor_mean_g)
)


kable(summary_fb_raw, caption = "Z' Factor per Run (fB)")
kable(summary_fb_avg, caption = "Z' Factor per Run (Average fB)")
kable(summary_g_raw, caption = "Z' Factor per Run (G)")
kable(summary_g_avg, caption = "Z' Factor per Run (Average G)")







```




# Scatterplot of S- & G- Coordinates of STA Treatments


```{r plot1, echo=FALSE, warning=FALSE, message=FALSE,  fig.width=13, fig.height=10, dpi=330}

library(ggplot2)
library(viridis)



# previously colored by Plate)ID (org1) but this is only 1 plate.

scatterplot <- ggplot(dataframe, aes(x=G, y=S, color =Experiment)) + geom_point(size=4) + facet_wrap(~Replicate_Group + Concentration) + labs(x = "G-coordinate", y = "S-coordinate", color = "Experiment") + scale_color_brewer(palette = "Dark2") + theme_bw(base_size=20) + ggtitle("S- & G- Coordinates") 


scatterplot


```




# Mixed Effects Models (Fraction Bound)


```{r boxplot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=9, dpi=300}

#setting up STA-1uM_df

library(dplyr)
library(ggplot2)
library(lme4)
library(ggbeeswarm)

df_all <- dataframe %>% mutate(Concentration = case_when(
  Replicate_Group == "untreated" ~ 0,
  Replicate_Group == "STA-0-1uM" ~ 0.1,
  Replicate_Group == "STA-1uM" ~ 1,
  TRUE ~ NA_real_
))


averages <- df_all %>%
  group_by(Experiment, Concentration) %>%
  summarize(mean_fB = mean(fB, na.rm=TRUE), .groups='drop')



box_superplot <- ggplot(df_all, aes(x = as.factor(Concentration), y = fB)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  geom_quasirandom(aes(color = as.factor(Experiment)), size = 2, alpha = 0.9) +
  geom_beeswarm(data = averages, aes(x = as.factor(Concentration), y = mean_fB,color = as.factor(Experiment)), size = 5)+ 
  scale_color_brewer(palette = "Dark2") +
  theme_bw(base_size=20) +
  ggtitle("Fraction Bound Across Conditions") +
  labs(x = "STA Concentration uM", y = "Fraction Bound", color = "Experiment")
  
  
box_superplot

```









```{r fixed, echo=FALSE, warning=FALSE, message=FALSE, fig.width=11, fig.height=9, dpi=300}




# Fixed effects model (not considering experiment)
#invisible(fixed_effects_model <- lm(fB ~ Concentration, data = df_all))


# Fixed effects plot
fixed_plot <- ggplot(data = df_all, aes(x = Concentration, y = fB)) +
    geom_point(alpha=0.9, color = "darkgray", size =3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    theme_bw(base_size=20) +
    ggtitle("STA Fraction Bound") + 
    labs(subtitle = "Fixed Effects Visualizer") +
    ylab("Fraction Bound") +
    xlab("STA Concentration (uM)")
  

fixed_plot

```







```{r mem, echo=FALSE, warning=FALSE, message=FALSE,fig.width=12, fig.height=9, dpi=300}
# Mixed effects model
#invisible(mixed_effects_model <- lmer(fB ~ Concentration  + (1 | Experiment), data=df_all))


mixed_effects_plot_w_date <- ggplot(data = df_all, aes(x = Concentration, y = fB, color = as.factor(Experiment))) +
  geom_point(alpha=0.9, size=3) +
    geom_smooth(method = "lm", aes(group=Experiment), se = FALSE, size =1.2) +
  theme_bw(base_size=20) +
  scale_color_brewer(palette = "Dark2") +
    ggtitle("STA Fraction Bound") + 
    labs(subtitle = "Mixed Effects Visualizer") +
    labs(x = "STA Concentration (uM)", y = "Fraction Bound", color = "Experiment")


mixed_effects_plot_w_date



```

The mixed effect model examines the effects of STA Concentration on fraction bound, with experiment as a random variable to assess differences between experiment runs.

