---
title: "FLIM Analysis Template"
output: word_document
date: "2024-11-13"
author: "Sevde Coban"
---

```{r setup, include=FALSE, echo=FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

# Purpose

FLIM collects photon excitation information from cell images to examine metabolic changes within biologic samples. This information is Fourier transformed to collect two parameters: S-coordinate and G-coordinate. These parameters are used to calculate fraction bound (fB), which is the fraction of bound to unbound NADH within the cells we’re looking at. Fraction bound can tell you if the cells are using more glycolysis (more free NADH) or oxidative phosphorylation (more NADH is bound to an enzyme).

The purpose of this analysis is to compare the differences in fraction bound (fB) between conditions between the treatments DMSO and Osimertinib.


# Z' Factor and % Cell Viability for Fraction Bound and G-Coordinates

```{r fb, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(here)

df <- readRDS(here("data", paste0("data_cleaning_output_", ".RDS")))
avg_df <- readRDS(here("data", paste0("data_cleaning_output_avg_", ".RDS")))

df <- df %>%  
  group_by(scan_date) %>%
  mutate(cv_fb = sd(f_b, na.rm = TRUE)/mean(f_b, na.rm = TRUE)*100) %>%
  #Osim is low control
  #DMSO is high control
  mutate(ctrl_0_mean_fb = mean(f_b[treatment == "Osimertinib"], na.rm = TRUE),
         ctrl_0_sd_fb = sd(f_b[treatment == "Osimertinib"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_fb = f_b - ctrl_0_mean_fb,
         #take the positive control mean
         ctrl_100_mean_fb = mean(f_b[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_minus_mean_fb = mean(signal_minus_ctrl_0_mean_fb[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_sd_fb = sd(f_b[treatment == "DMSO"], na.rm = TRUE),
         z_prime_factor_fb = 1 - (3*ctrl_100_sd_fb + 3*ctrl_0_sd_fb)/abs(ctrl_100_mean_fb - ctrl_0_mean_fb),
         plate_cv_fb = mean(cv_fb, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime <- df$z_prime_factor_fb
plate_cv <- df$plate_cv_fb
plate_id <- df$plate_id

summary_df <- data.frame(plate_id = plate_id, z_prime_factor = z_prime, percent_cv = plate_cv)
summary_df <- summary_df[1, , drop = FALSE]

library(knitr)

kable(summary_df, caption = "Fraction Bound Z'factor and %CV")
```




```{r mean_fb, echo=FALSE, warning=FALSE, message=FALSE}


avg_df <- avg_df %>%  
  group_by(scan_date) %>%
  mutate(cv_mean_fb_mean = sd(mean_f_b, na.rm = TRUE)/mean(mean_f_b, na.rm = TRUE)*100) %>%
  #Osim is low control
  #DMSO is high control
  mutate(ctrl_0_mean_fb_mean = mean(mean_f_b[treatment == "Osimertinib"], na.rm = TRUE),
         ctrl_0_sd_fb_mean = sd(mean_f_b[treatment == "Osimertinib"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_fb_mean = mean_f_b - ctrl_0_mean_fb_mean,
         #take the positive control mean
         ctrl_100_mean_fb_mean = mean(mean_f_b[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_minus_mean_fb_mean = mean(signal_minus_ctrl_0_mean_fb_mean[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_sd_fb_mean = sd(mean_f_b[treatment == "DMSO"], na.rm = TRUE),
         z_prime_factor_mean_fb = 1 - (3*ctrl_100_sd_fb_mean + 3*ctrl_0_sd_fb_mean)/abs(ctrl_100_mean_fb_mean - ctrl_0_mean_fb_mean),
         plate_cv_mean_fb = mean(cv_mean_fb_mean, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime_mean_f_b <- avg_df$z_prime_factor_mean_fb
plate_cv_mean_fb <- avg_df$plate_cv_mean_fb
plate_id_mean_fb <- avg_df$plate_id

summary_df2 <- data.frame(plate_id = plate_id_mean_fb, z_prime_factor = z_prime_mean_f_b, percent_cv = plate_cv_mean_fb)
summary_df2 <- summary_df2[1, , drop = FALSE]

library(knitr)

kable(summary_df2, caption = "Mean Fraction Bound Z'factor and %CV")







```



```{r gcoord, echo=FALSE, warning=FALSE, message=FALSE}


df <- df %>%  
  group_by(scan_date) %>%
  mutate(cv_g = sd(g, na.rm = TRUE)/mean(g, na.rm = TRUE)*100) %>%
  #Osim is low control
  #DMSO is high control
  mutate(ctrl_0_mean_g = mean(g[treatment == "Osimertinib"], na.rm = TRUE),
         ctrl_0_sd_g = sd(g[treatment == "Osimertinib"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_g = g - ctrl_0_mean_g,
         #take the positive control mean
         ctrl_100_mean_g = mean(g[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_minus_mean_g = mean(signal_minus_ctrl_0_mean_g[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_sd_g = sd(g[treatment == "DMSO"], na.rm = TRUE),
         z_prime_factor_g = 1 - (3*ctrl_100_sd_g + 3*ctrl_0_sd_g)/abs(ctrl_100_mean_g - ctrl_0_mean_g),
         plate_cv_g = mean(cv_g, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime <- df$z_prime_factor_g
plate_cv <- df$plate_cv_g
plate_id <- df$plate_id

summary_df2 <- data.frame(plate_id = plate_id, z_prime_factor = z_prime, percent_cv = plate_cv)
summary_df2 <- summary_df2[1, , drop = FALSE]

library(knitr)

kable(summary_df2, caption = "G-Coordinate Z'factor and %CV")







```


```{r mean gcoord, echo=FALSE, warning=FALSE, message=FALSE}

avg_df <- avg_df %>%  
  group_by(scan_date) %>%
  mutate(cv_mean_g_mean = sd(mean_g_coord, na.rm = TRUE)/mean(mean_g_coord, na.rm = TRUE)*100) %>%
  #Osim is low control
  #DMSO is high control
  mutate(ctrl_0_mean_g_mean = mean(mean_g_coord[treatment == "Osimertinib"], na.rm = TRUE),
         ctrl_0_sd_g_mean = sd(mean_g_coord[treatment == "Osimertinib"], na.rm = TRUE),
         #take signal minus the DMSO_mean
         #(this is divided by positive control to get our normalized signal)
         signal_minus_ctrl_0_mean_g_mean = mean_g_coord - ctrl_0_mean_g_mean,
         #take the positive control mean
         ctrl_100_mean_g_mean = mean(mean_g_coord[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_minus_mean_g_mean = mean(signal_minus_ctrl_0_mean_g_mean[treatment == "DMSO"], na.rm = TRUE),
         ctrl_100_sd_g_mean = sd(mean_g_coord[treatment == "DMSO"], na.rm = TRUE),
         z_prime_factor_mean_g = 1 - (3*ctrl_100_sd_g_mean + 3*ctrl_0_sd_g_mean)/abs(ctrl_100_mean_g_mean - ctrl_0_mean_g_mean),
         plate_cv_mean_g = mean(cv_mean_g_mean, na.rm = TRUE)) %>%
        mutate(dmso_percent = NA,
         starting_media = NA,
         day = NA,
         reader = NA) %>% ungroup()


z_prime_mean_g_coord <- avg_df$z_prime_factor_mean_g
plate_cv_mean_g_coord <- avg_df$plate_cv_mean_g
plate_id_mean_g <- avg_df$plate_id

summary_df2 <- data.frame(plate_id = plate_id_mean_g, z_prime_factor = z_prime_mean_g_coord, percent_cv = plate_cv_mean_g_coord)
summary_df2 <- summary_df2[1, , drop = FALSE]

library(knitr)

kable(summary_df2, caption = "Mean G-Coordinate Z'factor and %CV")









```

# Scatterplot of S- & G- Coordinates of Treatments DMSO, Osimertinib, and Temsirolimus


```{r plot, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=6}

library(ggplot2)
library(viridis)


scatterplot <- ggplot(df, aes(x=g, y=s, color = plate_id)) + geom_point(size=2) + facet_wrap(~treatment + concentration) + labs(x = "g-coordinate", y = "s-coordinate", color = "Plate_ID") + scale_color_manual(values = c("darkolivegreen4", "darksalmon", "lightblue")) + theme_bw() + ggtitle("S- & G- Coordinates") 


scatterplot







```


```{r same distribution, echo=FALSE, warning=FALSE, message=FALSE}

set.seed(200)

same_dist_df <- df

numeric_columns <- sapply(same_dist_df, is.numeric)
for (col in names(same_dist_df)[numeric_columns]) {
  mean_val <- mean(df[[col]])
  sd_val <- sd(df[[col]])
  same_dist_df[[col]] <- rnorm(nrow(df), mean = mean_val, sd = sd_val)
}

same_dist_df$plate_id <- "UP02"
same_dist_df <- same_dist_df[sample(nrow(same_dist_df)), ]



```





```{r different_dist, echo=FALSE, warning=FALSE, message=FALSE}

different_dist_df <- df

set.seed(200)

numeric_columns_diff <- sapply(different_dist_df, is.numeric)
for (col in names(different_dist_df)[numeric_columns]) {
  mean_val_diff <- mean(df[[col]])
  sd_val_diff <- sd(df[[col]])
  different_dist_df[[col]] <- rnorm(nrow(df), mean = mean_val_diff + 10, sd = sd_val_diff *2)
}

different_dist_df$plate_id <- "UP03"
different_dist_df <- different_dist_df[sample(nrow(different_dist_df)), ]


```

# Mixed Effects Models

## DMSO


```{r mem, echo=FALSE, warning=FALSE, message=FALSE}

#setting up dmso_df

library(dplyr)
library(ggplot2)
library(lme4)

df <- df %>% mutate(plate_id = "UP01")
same_dist_df <- same_dist_df %>% mutate(plate_id = "UP02")
different_dist_df <- different_dist_df %>% mutate(plate_id = "UP03")

combined_df <- bind_rows(df, same_dist_df, different_dist_df)

dmso_df <- combined_df %>% filter(treatment == "DMSO")

model <- lmer(f_b ~ concentration + (1 | plate_id), data = dmso_df)

summary(model)
  
plot <- ggplot(data = dmso_df, aes(x = concentration, y = f_b)) +
    geom_point(color = "gray") +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    theme_bw() +
    facet_wrap(~ cell_line) +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Fixed Effects Visualizer") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")
    
plot_w_date <- ggplot(data = dmso_df, aes(x = concentration, y = f_b)) +
    geom_point(aes(color = as.factor(plate_id))) +
    geom_smooth(aes(color = as.factor(plate_id), group = plate_id), method = "lm", se = FALSE) +
    theme_bw() +
    facet_wrap(~ cell_line) +
    scale_color_viridis_d() +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Mixed Effects Visualizer") +
    labs(color = "Scan Date") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")


plot
plot_w_date


```

## Osimertinib


```{r mem2, echo=FALSE, warning=FALSE, message=FALSE}

#setting up dmso_df

library(dplyr)
library(ggplot2)
library(lme4)

df <- df %>% mutate(plate_id = "UP01")
same_dist_df <- same_dist_df %>% mutate(plate_id = "UP02")
different_dist_df <- different_dist_df %>% mutate(plate_id = "UP03")

combined_df <- bind_rows(df, same_dist_df, different_dist_df)

os_df <- combined_df %>% filter(treatment == "Osimertinib")

model <- lmer(f_b ~ concentration + (1 | plate_id), data = dmso_df)

summary(model)
  
plot <- ggplot(data = dmso_df, aes(x = concentration, y = f_b)) +
    geom_point(color = "gray") +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    theme_bw() +
    facet_wrap(~ cell_line) +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Fixed Effects Visualizer") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")
    
plot_w_date <- ggplot(data = os_df, aes(x = concentration, y = f_b)) +
    geom_point(aes(color = as.factor(plate_id))) +
    geom_smooth(aes(color = as.factor(plate_id), group = plate_id), method = "lm", se = FALSE) +
    theme_bw() +
    facet_wrap(~ cell_line) +
    scale_color_viridis_d() +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Mixed Effects Visualizer") +
    labs(color = "Scan Date") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")


plot
plot_w_date



```
## Temsirolimus

```{r mem3, echo=FALSE, warning=FALSE, message=FALSE}

#setting up dmso_df

library(dplyr)
library(ggplot2)
library(lme4)

df <- df %>% mutate(plate_id = "UP01")
same_dist_df <- same_dist_df %>% mutate(plate_id = "UP02")
different_dist_df <- different_dist_df %>% mutate(plate_id = "UP03")

combined_df <- bind_rows(df, same_dist_df, different_dist_df)

tem_df <- combined_df %>% filter(treatment == "Temsirolimus")

model <- lmer(f_b ~ concentration + (1 | plate_id), data = dmso_df)

summary(model)
  
plot <- ggplot(data = tem_df, aes(x = concentration, y = f_b)) +
    geom_point(color = "gray") +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    theme_bw() +
    facet_wrap(~ cell_line) +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Fixed Effects Visualizer") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")
    
plot_w_date <- ggplot(data = dmso_df, aes(x = concentration, y = f_b)) +
    geom_point(aes(color = as.factor(plate_id))) +
    geom_smooth(aes(color = as.factor(plate_id), group = plate_id), method = "lm", se = FALSE) +
    theme_bw() +
    facet_wrap(~ cell_line) +
    scale_color_viridis_d() +
    ggtitle("Fraction Bound") + 
    labs(subtitle = "Mixed Effects Visualizer") +
    labs(color = "Scan Date") +
    ylab("Fraction Bound (f_b)") +
    xlab("Concentration (uM)")


plot
plot_w_date



```


